Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-28 LICS21/antitone.tex
Minor

For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
defined as: for $n \in \mathbb N$
\begin{align*}
{\Theta_2} \, n 
&\mapsto 
g_2(n)
\\
0 &\mapsto 0.
\end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}
& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
  &=  \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
  &=  \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
  &>  \frac 1 {4n(n+2)} \\
  &>  \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
\end{align*}
\changed[lo]{(The first inequality follows from the fact that for all $x \in (0, 1)$, $\log (1 + x) > x$.)}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.

@@

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
\, 10
\]
To construct an antitone ranking function, we solve the recurrence relation:
% \[
% \left\{
% \begin{array}{rll}
% z_0 &=& 0\\
% n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
% \end{array}
% \right.
% \]
\begin{align*}
z_0 &= 0\\
n \geq 1, \quad z_n &> \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
\end{align*}
For $n > 2$, $z_n = \log(n-1)$ works. The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$,\lo{The preceding inequality is incorrect: for $x \in (0, 1)$, $\log (1 + x) > x$.} this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone. For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$. Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.

>>>>

For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
defined as: for $n \in \mathbb N$
\begin{align*}
{\Theta_2} \, n 
&\mapsto 
g_2(n)
\\
0 &\mapsto 0.
\end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}
& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
  &=  \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
  &=  \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
  &>  \frac 1 {4n(n+2)} \\
  &>  \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
\end{align*}
\changed[lo]{(The first inequality follows from the fact that for all $x \in (0, 1)$, $\log_2 (1 + x) > x$.)}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.

@@

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
\, 10
\]
To construct an antitone ranking function, we solve the recurrence relation:
% \[
% \left\{
% \begin{array}{rll}
% z_0 &=& 0\\
% n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
% \end{array}
% \right.
% \]
\begin{align*}
z_0 &= 0\\
n \geq 1, \quad z_n &> \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
\end{align*}
For $n > 2$, $z_n = \log(n-1)$ works. The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$,\lo{The preceding inequality is incorrect: for $x \in (0, 1)$, $\log (1 + x) > x$.}
\lo{OK. You mean $\ln$ rather than $\log_2$.} this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone. For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$. Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.
